---
title       : Leveraged ETF Returns Simulator
subtitle    : Coursera, Data Science, Developing Data Products, Course Project
author      : Patrick Charles
job         : interval software engineering
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides

--- .class #id 

## Leveraged ETFs - Intro

### Background

* ETF - exchanged traded funds
* Leverage - derivatives and debt to magnify returns of an underlying index
* Funds typically target multipliers of 2x or 3x to amplify daily returns
* Long-term performance of leveraged ETFs is not well understood

### Goal

Use simulation, analysis and visualization to understand the effects of
leverage and long-term performance characteristics.

R is used to simulate different leverage scenarios. Shiny is 
be used to allow users to interact with, understand
and visualize the long-term effect on performance of different
leverage multipliers.

--- .class #id

## Leveraged ETF - Returns

### Why are these interesting?

 * Leverage is applied daily, e.g. for 2x.. 
   * A 5% increase in the underlying index causes a 10% gain for the ETF
   * A 5% decrease in the underlying index causes a 10% loss for the ETF

 * Straightforward, but over time: compounding effects and price decay due to volatility
   * e.g. take series of gains and losses: (1) 5%, (2) -10%, (3) -5%, (4) +10%
   * underlying index @ $100 -> $105 -> $94.50 -> $89.78 -> $98.75
   * 2x leveraged.. $100 -> $110 -> $88 -> $79.20 -> $95.04

--- .class #id 

## Leveraged ETF - Underlying Index

### R can be used to load historical closing price data, e.g. NASDAQ 

```{r load_index, message=FALSE, warning=FALSE, echo=FALSE, fig.width=10, fig.height=5}
  library(ggplot2)
  library(quantmod)
  xts.daily <- getSymbols(Symbols="^IXIC", src="yahoo",
                          from="2005-01-01", to="2015-06-01",
                          env=NULL)
  df.daily <- as.data.frame(xts.daily)
  colnames(df.daily) <- c("Open", "High", "Low", "Actual", "Volume")
  df.daily$Date <- as.Date(rownames(df.daily)) # give date a column
  ggplot(df.daily, aes(x=Date, y=Actual)) + 
         geom_line() + 
         xlab("Date") + ylab("Closing Price") +
         ggtitle("NASDAQ Composite Daily Closing Price (Actual)")
```

--- .class #id 

## Leveraged ETF - Simulation

### R can be used to simulate leveraged daily returns
* by calculating price percent deltas based on historical closing prices
* applying the deltas (with leverage) to simulate daily ETF returns
* visualize cumulative leveraged vs. underlying (non-leveraged) returns over time

```{r simulate, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=5}
  library(reshape2)

  leverage <- 2.0
  df.daily$Delta <- Delt(df.daily$Actual)[, 1] # daily price deltas
  df.daily$Simulated <- df.daily$Actual[1] # seed the leverage col

  # calculate daily leveraged returns
  for(i in 2:nrow(df.daily)) { 
    previous <- df.daily[i - 1, ]$Simulated # previous value
    delta <- df.daily[i, ]$Delta # closing price delta
    df.daily[i, ]$Simulated = # calculated, next leveraged price
    previous + previous * delta * leverage
  }

  # render a plot of actual vs. simulated returns
  df.melted <- melt(df.daily, id.vars=c("Date"))
  df.melted <- df.melted[df.melted$variable == 'Actual' |
                         df.melted$variable == 'Simulated', ]
  colnames(df.melted) <- c("Date", "Price", "Value")
  ggplot(df.melted, aes(x=Date, y=Value, color=Price)) +
    geom_line() + 
    xlab("Date") + ylab("Closing Price") +
    ggtitle("NASDAQ Actual vs. 2x Leveraged")
```

--- .class #id

## Leveraged ETF - Interactive Simulation

### As a Shiny Application..

* Allows a user to...
  * Select from a variety of underlying securities
  * Load historical data for a date range of interest
  * Interactively modify the ETF leverage multipliers
  * Instantly see the simulated performance of leveraged vs. non-leveraged funds

### Try It!
* shinyapps.io: [etf-leverage-simulator](https://pchuck.shinyapps.io/etf-leverage-simulator)

### The Source Code
* github.com: [pchuck/etf-leverage-simulator](https://github.com/pchuck/etf-leverage-simulator)

